---
title: "STAT_400-MLM_Milestone1"
author: "Ishan Agrahar, Jake Sturges, Noah Tobias, Prajwal Bhandari"
date: "2025-12-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Front Matter

```{r}
# Libraries
library(tidyverse)

# Dataset Read-In
testScores <- read.csv("data/test_scores.csv")
```


## Dataset Choice

The dataset that our group chose for the MLM portion of the project is Dataset 1 (Education Related).

This dataset is from Kaggle and is titled *Predict Test Scores of students*.

You may find the dataset online at the following link: <https://www.kaggle.com/datasets/kwadwoofosu/predict-test-scores-of-students/data>

```{r}
glimpse(testScores)
```

The Kaggle page already has descriptions for the data so I shall simply copy paste it here for reference.

`school` - Name of the school the student is enrolled in - Categorical, 23 Levels

`school_setting` - The location of the school - Categorical, 3 Levels

`school_type` - The type of school. Either public or non-public - Categorical, 2 Levels

`classroom` - The type of classroom - Categorical, 97 Levels

`teaching_method` - Teaching methods: Either experimental or Standard - Categorical, 2 Levels

`n_student` - Number of students in the class - Quantitative

`student_id` - A unique ID for each student - Categorical, 2133 Levels

`gender` - The gender of the students: male or female - Categorical, 2 Levels

`lunch` - Whether a student qualifies for free/subsidized lunch or not - Categorical, 2 Levels

`pretest` - The pretest score of the students out of 100 - Quantitative

`posttest` - The posttest scores of the students out of 100 - Quantitative

## Research Question

The research question that our group asks is what is the association between a schools setting and a schools teaching method?

## Response

## Level 1 Observational Units



## Level 2 Observational Units

## Level 1 Variables

## Level 2 Variables

## Fixed Effects

## Random Effects